Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-02 abstract.tex
tweaks and comments in intro/abstract
bla
intro and other stuff

%!TEX root =  matrix-est.tex
We propose an efficient procedure for estimating the mixing time of a Markov chain.
The novelty of our approach lies in several key aspects.
Crucially,
we do not assume any a priori bounds
on the mixing time,
which
sets our approach apart from
most others that do
require some advance knowledge of this quantity.
Another
advantage of our result over
common practice is the sufficiency of a single long sampling sequence,
without access to a restart mechanism.
Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
finite-sample deviation bounds without any knowledge of the mixing coefficients.
Upper and lower sample complexity bounds are provided and challenging open problems are posed.

>>>>

%!TEX root =  matrix-est.tex
This article provides an efficient procedure for estimating the mixing
time of a reversible ergodic Markov chain.
The procedure computes the estimate from a single sample path of the
Markov chain, and does not require any \emph{a priori} bounds on the
mixing time itself.
This stands in contrast to previous approaches, which either require a
reset mechanism or require additional prior knowledge.
Most notably, the procedure produces a fully empirical confidence
interval (i.e., it depends only on the sample path and a given
confidence parameter $\delta$) that is always valid with probability
$1-\delta$; and the width of the interval converges to zero at a
$\sqrt{n}$-rate, where $n$ is the length of the sample path.

%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
%finite-sample deviation bounds without any knowledge of the mixing coefficients.
%
%
%The novelty of our approach lies in several key aspects.
%Crucially,
%we do not assume any a priori bounds
%on the mixing time,
%which
%sets our approach apart from
%most others that do
%require some advance knowledge of this quantity.
%Another
%advantage of our result over
%common practice is the sufficiency of a single long sampling sequence,
%without access to a restart mechanism.
%Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
%estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
%finite-sample deviation bounds without any knowledge of the mixing coefficients.
%%Upper and lower sample complexity bounds are provided and challenging open problems are posed.
%%\todod{I would remove the last sentence, at least the part about
%%``challenging'' open problems.}
%%%test

### 2015-06-02 empirical.tex
intro and other stuff

The main result of this section is the following result:
\begin{theorem}
  \label{thm:empirical}
  Suppose \cref{alg:empest} is given as input a sample path of length
  $n$ from an ergodic and reversible Markov chain and confidence
  parameter $\delta \in (0,1)$.
  Let $\gap>0$ denote the spectral gap, $\vpi$ the unique stationary
  distribution, and $\pimin>0$ the minimum stationary probability.
  Then, on an event of probability at least $1-\delta$,
  \[
    \pi_i \in [\hat\pi_i-\hat{b},\hat\pi_i+\hat{b}]
    \quad \text{for all $i \in [d]$} ,
    \qquad\text{and}\qquad
    \gap \in [\hatgap-\hat{w},\hatgap+\hat{w}]
    .
  \]
  Moreover, $\hat{b}$ and $\hat{w}$ almost surely satisfy (as $n \to
  \infty$)
  \[
    \hat{b}
    =
    O\Parens{
      \max_{(i,j) \in [d]^2}
      \kappa
      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
    }
    ,
    \quad
    \hat{w}
    =
     O\Parens{ 
     \frac{\kappa}{\pimin} \sqrt{\frac{\ln(d/\delta)}{\pimin n}} + 
      \sqrt{ \frac{d\ln(d/\delta)}{ \pimin n}}
    }
    ,
  \]
  where $\kappa$ is analogous to the $\hat{\kappa}$ computed
  by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
  $\giA$ of $\vA := \vI - \vP$.
  \todoc{
    Precisely,
    what we mean by the $O(\cdot)$ notation is as follows:
  for a random sequence of nonnegative numbers $(X_n)_n$
  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
  indexed by $\eps$ we
  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
  if for some universal constant $C>0$, 
  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
  the choice of $\eps$. I think this should be made clear.
  }
\end{theorem}
The proof of \cref{thm:empirical} is given in \cref{app:empirical}.
As mentioned above, the obstacle encountered in \cref{thm:err} is
avoided by exploiting the Markov property.
We establish fully observable upper and lower bounds on the entries of
$\vP$ that converge at a $\sqrt{n}$ rate using standard martingale
tail inequalities; this justifies the validity of the bounds from
Step~\ref{step:P-bound}.
Properties of the group inverse~\citep{meyer1975role,cho2001comparison} and eigenvalue
perturbation theory~\citep{stewart1990matrix} are used to validate the
empirical bounds on $\pi_i$ and $\gap$ developed in the remaining
steps of the algorithm.

>>>>

The main result of this section is the following result:
\begin{theorem}
  \label{thm:empirical}
  Suppose \cref{alg:empest} is given as input a sample path of length
  $n$ from an ergodic and reversible Markov chain and confidence
  parameter $\delta \in (0,1)$.
  Let $\gap>0$ denote the spectral gap, $\vpi$ the unique stationary
  distribution, and $\pimin>0$ the minimum stationary probability.
  Then, on an event of probability at least $1-\delta$,
  \[
    \pi_i \in [\hat\pi_i-\hat{b},\hat\pi_i+\hat{b}]
    \quad \text{for all $i \in [d]$} ,
    \qquad\text{and}\qquad
    \gap \in [\hatgap-\hat{w},\hatgap+\hat{w}]
    .
  \]
  Moreover, $\hat{b}$ and $\hat{w}$ almost surely satisfy (as $n \to
  \infty$)
  \[
    \hat{b}
    =
    O\Parens{
      \max_{(i,j) \in [d]^2}
      \kappa
      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
    }
    ,
    \quad
    \hat{w}
    =
     O\Parens{ 
     \frac{\kappa}{\pimin} \sqrt{\frac{\ln(d/\delta)}{\pimin n}} + 
      \sqrt{ \frac{d\ln(d/\delta)}{ \pimin n}}
    }
    ,
  \]
  where $\kappa$ is analogous to the $\hat{\kappa}$ computed
  by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
  $\giA$ of $\vA := \vI - \vP$.\footnote{%
    In \cref{thm:empirical,thm:combined}, our use of big-$O$ notation
    is as follows.
    For a random sequence $(Y_n)_n$ and a (non-random) positive
    sequence $(a_{\theta,n})_n$ parameterized by $\theta$, we say
    ``$Y_n = O(a_{\theta,n})$ holds almost surely as $n\to\infty$'' if
    there is some universal constant $C>0$ such that for all $\theta$,
    $\limsup_{n\to\infty} Y_n/a_{\theta,n} \leq C$ holds almost
    surely.%
  }%
%  \todoc{
%    Precisely,
%    what we mean by the $O(\cdot)$ notation is as follows:
%  for a random sequence of nonnegative numbers $(X_n)_n$
%  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
%  indexed by $\eps$ we
%  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
%  if for some universal constant $C>0$, 
%  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
%  the choice of $\eps$. I think this should be made clear.
%  }
\end{theorem}
The proof of \cref{thm:empirical} is given in \cref{app:empirical}.
As mentioned above, the obstacle encountered in \cref{thm:err} is
avoided by exploiting the Markov property.
We establish fully observable upper and lower bounds on the entries of
$\vP$ that converge at a $\sqrt{n}$ rate using standard martingale
tail inequalities; this justifies the validity of the bounds from
Step~\ref{step:P-bound}.
Properties of the group inverse~\citep{meyer1975role,cho2001comparison} and eigenvalue
perturbation theory~\citep{stewart1990matrix} are used to validate the
empirical bounds on $\pi_i$ and $\gap$ developed in the remaining
steps of the algorithm.
